Effects of Reynolds Number on the Energy Conversion and Near-Wake Dynamics of a High Solidity Vertical-Axis Cross-Flow Turbine

### 2016-01-20 paper.tex
Suggested changes from editor

\documentclass[energies,article,accept,moreauthors,pdftex,12pt,a4paper]{mdpi}

%=================================================================
\setcounter{page}{1}
\lastpage{x}
\doinum{10.3390/------}
\pubvolume{xx}
\pubyear{2015}
\history{Received: xx / Accepted: xx / Published: xx}
%------------------------------------------------------------------
% The following line should be uncommented if the LaTeX file is uploaded to
% arXiv.org
%\pdfoutput=1

@@

% Add packages and commands to include here
% The amsmath, amsthm, amssymb, hyperref, caption, float and color packages are
% loaded by the MDPI class.
\usepackage{graphicx}
%\usepackage{subfigure,psfig}
\usepackage[draft]{todonotes}
\usepackage{lineno}
\usepackage[utf8]{inputenc}

\def \p{\partial}
\def \d{\mathrm{d}}
\def \D{\mathrm{D}}

%=================================================================
%% Please use the following mathematics environments:
%\theoremstyle{mdpi}
%\newcounter{thm}
%\setcounter{thm}{0}
%\newcounter{ex}
%\setcounter{ex}{0}
%\newcounter{re}
%\setcounter{re}{0}
%\newtheorem{Theorem}[thm]{Theorem}
%\newtheorem{Lemma}[thm]{Lemma}
%\newtheorem{Characterization}[thm]{Characterization}
%\newtheorem{Proposition}[thm]{Proposition}
%\newtheorem{Property}[thm]{Property}
%\newtheorem{Problem}[thm]{Problem}
%\newtheorem{Example}[ex]{Example}
%\newtheorem{Remark}[re]{Remark}
%\newtheorem{Corollary}[thm]{Corollary}
%\newtheorem{Definition}[thm]{Definition}
%% For proofs, please use the proof environment (the amsthm package is loaded by
% the MDPI class).

@@

% Authors (Add full first names)
\Author{Peter Bachant $^{1,}$* and Martin Wosnik $^{1}$*}

% Affiliations / Addresses (Add [1] after \address if there is only one
% affiliation.)
\address{%
$^{1}$ Center for Ocean Renewable Energy, University of New Hampshire, 24
Colovos Rd., Durham, NH, USA}

% Contact information of the corresponding author (Add [2] after \corres if
% there are more than one corresponding author.)
\corres[2]{pxL3@unh.edu, martin.wosnik@unh.edu}

\linenumbers

% Abstract (Less than 200 words)
\abstract{Experiments were performed with a large laboratory-scale high solidity
    cross-flow turbine to investigate Reynolds number effects on performance and
    wake characteristics, and to establish scale thresholds for physical and
    numerical modeling of individual devices and arrays. It was demonstrated that
    the performance of the cross-flow turbine becomes essentially $Re$-independent
    at a Reynolds number based on turbine diameter of $Re_D \approx 10^6$, or an
    approximate average Reynolds number based on blade chord of $Re_c \approx 2
    \times 10^5$. A simple model that calculates peak torque coefficient from static
    foil data and cross-flow turbine kinematics was shown to be a reasonable
    estimator for predicting Reynolds number dependence of an actual cross-flow
    turbine operating under dynamic conditions. Mean velocity and turbulence
    measurements in the near-wake showed subtle differences over the range of $Re$
    investigated. However, when transport terms for the streamwise momentum and mean
    kinetic energy were calculated, a similar $Re$ threshold was revealed. These
    results imply that physical model studies of cross-flow turbines should achieve
    $Re_D \sim 10^6$ to properly approximate both the performance and wake dynamics
    of full scale devices and arrays.}

% Keywords: add 3 to 10 keywords
\keyword{Reynolds number; cross-flow turbine; turbine performance; marine
hydrokinetic energy; wind energy; VAWT; scale model}

@@

Scaled physical models are often used in science and engineering to approximate
real-world systems. The principle of dynamic similarity allows for geometrically
scaled systems to be dynamically identical if certain nondimensional physical
parameters are matched. In the case of fluid systems, the most important
nondimensional parameter is often the Reynolds number, $Re$, which quantifies
the importance of inertial forces over viscous forces on the flow physics
\cite{Acheson1990}: $Re \equiv Ul/\nu$, where $U$ and $l$ are characteristic
velocity and length scales, respectively, and $\nu$ is the fluid kinematic
viscosity. The other common dynamical scale---for systems with a free
surface---is the Froude number $Fr \equiv U/\sqrt{gl}$, where $g$ is the
gravitational acceleration. Matching both Froude and Reynolds number is not
possible for a scaled model in a given fluid, which is illustrated by the
relation $Re = l^{3/2} g^{1/2} Fr / \nu$, since $Re$ scales linearly with $l$,
or the geometric scale. In general the geometric scale of a physical model is
not necessarily the same as its dynamical scale. As such, hereafter we will use
the word ``scale'' to refer to this dynamical scale or Reynolds number and
assume Froude number is small enough to be negligible.

With regards to wind and marine hydrokinetic (MHK) turbines, scaled physical
models are used to validate predictive numerical models, predict full-scale
performance of individual turbines, and design or investigate arrays of devices.
Scaled models have the benefit of being significantly less expensive, however, a
key question is whether or not an acceptable scale mismatch exists. Similarly,
it is questionable whether numerical models validated with physical model data
obtained orders of magnitude away from prototype scale should be considered
validated at all. An example of the errors that can result from the
extrapolation of small scale experiments can be found in \cite{Baker1991}.

The performance and wake characteristics of cross-flow (vertical-axis) turbines
depend on turbine solidity, blade profile (lift/drag, dynamic stall at reduced
frequency of turbine rotation, symmetry, thickness, camber), blade pitch, number
of blades, strut drag and operational parameters such as tip speed ratio---{\em
    and} on Reynolds number \cite{Para2002}. Note that an average blade chord
Reynolds number, $Re_{c,\mathrm{avg}} \approx \lambda U_\infty c/ \nu$, can be
expressed in terms of tip speed ratio $\lambda \equiv \omega R/ U_\infty$, where
$U_\infty$ is the free stream velocity, $c$ is the blade chord length, $\nu$ is
the fluid kinematic viscosity, $\omega$ is the rotor's angular velocity, and $R$
is the rotor radius. The value of $\lambda$ at which a turbine reaches peak
performance in general decreases with turbine solidity $Nc/(\pi D)$
\cite{Templin1974}, which allows for the use of a simpler Reynolds number based
on turbine diameter $Re_D \equiv U_\infty D/\nu$.

Solidity often directly correlates with the chord-to-radius ratio $c/R$. Rotors
with $c/R \ge 0.1$ are considered high solidity \cite{Fiedler2009}, for which
so-called flow curvature or virtual camber effects become significant
\cite{Migliore1980}. These effects arise from the blade sections' circular
paths, and can complicate the comparison with behavior of an airfoil in a linear
flow.

Blackwell et al. \cite{Blackwell1976} investigated the effects of Reynolds
number on the performance of a 2 m diameter Darrieus vertical-axis cross-flow
wind turbine with NACA 0012 blades. By varying wind tunnel speed, the turbine
was made to operate at approximately constant blade chord Reynolds number $Re_c$
ranging from $1 \times 10^5$ to $3 \times 10^5$. In this regime the turbine
power coefficient $C_P$ was shown to be sensitive to $Re_c$, with sensitivity
diminishing at the higher Reynolds numbers, especially for turbines of lower
solidity ($Nc/R$, where $N$ is the number of blades and $R$ is the turbine's
maximum radius). More recently, Polagye and Cavagnaro \cite{Polagye2013b}
observed significant Reynolds number sensitivity for a high solidity helical
cross-flow turbine, and Bravo et al. \cite{Bravo2007} observed the power
coefficient of a straight-bladed VAWT to become Reynolds number independent at
$Re_c \approx 4 \times 10^5$. Bachant and Wosnik \cite{Bachant2014} observed
Reynolds number independence of power coefficient at optimal tip speed ratio for
a high solidity cross-flow turbine at $Re_D \approx 10^6$ or
$Re_{c,\mathrm{avg}} \approx 2 \times 10^5$.

The wake of a 2-D cross-flow turbine in the dynamic stall regime has been
studied via laser Doppler velocimetry by Brochier et al. \cite{Brochier1986} and
in 3-D with particle image velocimetry (PIV) by Fujisawa and Shibuya
\cite{Fujisawa2001}. However, both of these studies were performed at very low
Reynolds numbers---$Re_D = 10^4$ and $10^3$, respectively. Tescione et al.
\cite{Tescione2014} studied the wake of a vertical-axis wind turbine at optimal
tip speed ratio in a wind tunnel using stereo PIV at an approximate blade chord
Reynolds number $Re_c = 1.7 \times 10^5$; very close to the $Re$-independence
criteria reported in \cite{Bachant2014}, though it was not confirmed if this was
indeed a $Re$-independent state. The two lower $Re$ experiments tended to focus
on the effects of dynamic stall, whereas the higher $Re$ experiment mostly
concerned the mean velocity and tip vortex effects. Whereas the value of these
studies was in elucidating the complex wake dynamics of cross-flow turbines,
they also motivated the more systematic investigation of scale effects
undertaken here.

Scale effects on axial-flow or horizontal-axis wind turbines are better
understood, and investigators have methods for compensating. Krogstad and
Adaramola \cite{Krogstad2012a} observed Reynolds number independence of the
performance of a 0.9 m diameter axial-flow turbine rotor at $Re_D \approx 5
\times 10^5$ in wind tunnel tests. Their turbine blades had an S826 profile
along their entire span, a profile chosen for its $Re$-independence. Walker et
al. \cite{Walker2014} similarly chose a NACA 63-618 foil for their axial-flow
turbine tests in a towing tank, which were performed at a mid-span Reynolds
number $Re_c = 4 \times 10^5$. McTavish et al. \cite{McTavish2013} showed how
the near-wake expansion for an axial-flow rotor is increased at higher Reynolds
numbers, concluding that physical models should be designed with
$Re$-independence in mind if they are to be run at low $Re$. It is uncertain,
however, if these methods work equally well for cross-flow turbines.

When designing or studying arrays, it is common to use very small
(geometrically) scaled devices \cite{Chamorro2011, Chamorro2011b}. It is
therefore important to realize the limitations of evaluating both the power
output of turbines and the wake recovery when Reynolds number is very far from
full scale. Sometimes models used are not turbines, but wake generating objects,
e.g., porous disks, that are meant to replicate the wakes of real turbines
\cite{Goldenberg1983}. In this case, it is of interest to determine at what
Reynolds number one might be able to realistically study wake flows in an array,
and also to evaluate the effectiveness of a wake generator. In other words, a
wake generator may do a fine job simulating a scaled turbine, but how well can
it simulate a full-scale device? Note that for a porous disk, the Reynolds
number of interest is diameter-based, since it is the scaling of the far-wake
dynamics that matters.

Vertical-axis wind turbine array field experiments have revealed that improved
wake recovery allows for closer spacing when compared to conventional axial-flow
propeller-type turbines \cite{Dabiri2011, Kinzel2012}. It was observed
experimentally that a high solidity vertical-axis cross-flow turbine's near-wake
produces a unique vertical mean velocity field---generated by blade tip vortex
shedding---the advection by which is the largest contributor to streamwise
momentum and energy transport or recovery \cite{Bachant2015-JoT}. In this study
we seek to replicate those same momentum and energy balance considerations at
multiple Reynolds numbers, to examine the implications on how scaled, i.e., low
Reynolds number experiments may be used to study flows in turbine arrays.

@@

We are looking at how Reynolds number scaling affects both the blade loading,
i.e., turbine performance, and the near-wake. Typically, static airfoil data
shows that the static stall angle increases with blade chord Reynolds number
$Re_c$ \cite{Jacobs1937}. A review of Reynolds number effects on airfoil
behavior is presented in \cite{Lissaman1983}. In general, airfoil
performance---often characterized by the profile's lift-to-drag ratio---is
enhanced as the boundary layer on the foil transitions to turbulence closer to
the leading edge, which enables it to advance further downstream against the
adverse pressure gradient on the suction side, delaying separation to higher
angles of attack. For smooth airfoils, this transition can cause a dramatic
increase in foil performance at a blade chord Reynolds number on the order of
$10^5$ \cite{McMasters1980}. Note that there is a distinct lack of highly
reliable data for airfoils in this transitional regime and below. An evaluation
of the various databases relevant to cross-flow turbines is presented in
\cite{Bedon2014}.

Static foil performance does not tell the whole story for a cross-flow turbine.
The azimuthal, and therefore temporal variation of $\alpha$ in a cross-flow
turbine implies dynamic loading, encountering dynamic stall for tip speed ratios
near and below those of maximum rotor torque \cite{Para2002}. Bousman
\cite{Bousman2000-evaluation} states that dynamic stall is relatively
insensitive to Reynolds number for $Re=1.0 \times 10^5$--$2.5 \times
10^5$---judging from measurements on a pitching VR-7 foil in a wind and water
tunnel---since the loading is dominated by vortex shedding. However, Singleton
and Yeager Jr. \cite{Singleton2000} state that the effect of Reynolds number on
dynamic stall remains an unsolved question.

Despite lower lift on the blades at lower $Re$, we expect stronger tip vortex
shedding \cite{Yoon2005}. As mentioned previously, the dynamic stall vortex
shedding is not expected to be highly sensitive to Reynolds number, though a
larger separation bubble at lower $Re$ may induce higher levels of turbulence as
shed vortices become unstable and break down.

Chamorro et al. \cite{Chamorro2012} showed that turbulence statistics in an AFT
wake became $Re$-independent at $Re_D \approx 1 \times 10^5$, and that mean
velocity profiles became $Re$-independent slightly earlier at $Re_D \approx 5
\times 10^4$. Note that in this study the tip chord Reynolds number was not
reported, but is estimated to be $Re_c \approx 4 \times 10^4$ at $Re_D=1 \times
10^5$. It is therefore our objective to observe similar scaling relationships
for a cross-flow turbine near-wake.

@@

Experiments were performed in a turbine test bed specifically designed for
cross-flow turbines. The test bed was integrated as part of the UNH tow tank,
which is a 36 m long facility with a 3.66 m wide and 2.44 m deep cross-section.
The turbine model used in this study was the UNH Reference Vertical Axis Turbine
(RVAT), which was designed to be a generic case for numerical model testing,
similar to the Sandia National Labs/US Department of Energy RM2 River Turbine
\cite{Neary2014}, but with higher solidity, or blade chord-to-radius ratio.

The UNH-RVAT turbine has three blades made from NACA 0020 profiles with 0.14 m
chord. The blades are mounted at mid-chord w.r.t. the turbine axis, and the
turbine has a height (blade span) of 1 m and a diameter of 1 m, c.f.
Figure~\ref{fig:turbine}. The blockage ratio produced by the rotor's frontal
area is 11\%, which we decided not to correct for, as reliable methods are
not yet agreed upon for CFTs \cite{Cavagnaro2014}. As such, blockage should be
taken into account when using these data for model validation, i.e., the
experimental domain should be mimicked. The rotor has a relatively high solidity
$Nc/(\pi D) = 0.13$ and large chord-to-radius ratio $c/R = 0.28$. A CAD model of
the turbine is available from \cite{Bachant2014-RVAT-CAD}.

@@

\caption{UNH-RVAT turbine model. Turbine blades and struts made from NACA 0020
    profiles with 0.140 m chord. Note that the upper and lower mounting flanges have
    been excluded as these were included in the tare drag measurements.}

@@

The turbine was mounted in a frame constructed from NACA 0020 sections, shown in
Figure~\ref{fig:exp-setup}. The turbine shaft ran up through the water surface,
coupled to a Kollmorgen AKM permanent magnet servo motor with a 20:1 gearbox,
providing precise control over shaft RPM. This servo was controlled by the tow
tank's main motion controller for high synchronization with the carriage motion,
thereby giving precise measurement and control of tip speed ratio. An Interface
T8 200 Nm capacity rotary torque transducer was installed inline between the
servo and the turbine, and the servo was also mounted on a slewing ring bearing,
which allowed a redundant measurement of torque via an arm and load cell used to
counteract the turbine moment. The frame was mounted to the carriage via linear
guides, such that the total streamwise drag force was transferred to a pair of
Sentran ZB3 500 lbf capacity S-beam load cells, providing the rotor drag
measurements, after a separately measured tare drag was subtracted in post
processing. Similarly, a tare torque was measured by rotating the turbine shaft
in air. Turbine angular and tow carriage linear position were measured using
quadrature encoder signals, with $10^5$ counts-per-rev for the turbine and 10
$\mathrm{\mu m}$ resolution for the carriage position. These signals, along with
the torque and drag signals, were sampled at 2 kHz.

    \includegraphics[width=0.8\textwidth]{figures/exp-setup-photo}

    \includegraphics[width=0.8\textwidth]{figures/exp_setup_drawing}

    \caption{Experimental setup: turbine test bed installed in UNH tow tank.}

@@

Wake velocity was measured using a Nortek Vectrino+ acoustic Doppler velocimeter
(ADV), which has an approximately 6 mm diameter sampling volume, and sampled at
200 Hz. The probe was mounted on an automated positioning system, also
controlled by the tow tank's main motion controller. The ADV and DAQ systems'
sampling times were synchronized by triggering the start of data acquisition via
a pulse sent from the motion controller. Additional details of the turbine and
experimental setup are described in \cite{Bachant2015-JoT}.

@@

>>>>

\documentclass[energies,article,accept,moreauthors,pdftex,10pt,a4paper]{mdpi}

%=================================================================
\firstpage{1} %%MDPI internal note: new layout%%
\lastpage{\pageref*{LastPage}}
\makeatletter %%MDPI internal note: new layout%%
\setcounter{page}{\@firstpage} %%MDPI internal note: new layout%%
\makeatother %%MDPI internal note: new layout%%
\articlenumber{0000}
\doinum{10.3390/------}
\pubvolume{9}
\pubyear{2016}
\externaleditor{Academic Editor: Sukanta Basu}
\history{Received: 1 November 2015; Accepted: 15 January 2016; Published: date}
%------------------------------------------------------------------
% The following line should be uncommented if the LaTeX file is uploaded to
% arXiv.org
%\pdfoutput=1

@@

% Add packages and commands to include here
% The amsmath, amsthm, amssymb, hyperref, caption, float and color packages are
% loaded by the MDPI class.
\usepackage{graphicx}
%\usepackage{subfigure,psfig}
\usepackage[draft]{todonotes}
\usepackage{lineno}
\usepackage[utf8]{inputenc}
\usepackage{booktabs} 
\usepackage{multirow}
\usepackage{soul}
\usepackage{microtype}
\def \p{\partial}
\def \d{\mathrm{d}}
\def \D{\mathrm{D}}
\setitemize{parsep=6pt,itemsep=0pt,leftmargin=*,labelsep=5mm}
\setenumerate{parsep=6pt,itemsep=0pt,leftmargin=*,labelsep=5mm}
\setlist[description]{itemsep=0mm}

%=================================================================
%% Please use the following mathematics environments:
\theoremstyle{mdpi}
\newcounter{thm}
\setcounter{thm}{0}
\newcounter{ex}
\setcounter{ex}{0}
\newcounter{re}
\setcounter{re}{0}
\newtheorem{Theorem}[thm]{Theorem}
\newtheorem{Lemma}[thm]{Lemma}
\newtheorem{Characterization}[thm]{Characterization}
\newtheorem{Proposition}[thm]{Proposition}
\newtheorem{Property}[thm]{Property}
\newtheorem{Problem}[thm]{Problem}
\newtheorem{Example}[ex]{Example}
\newtheorem{Remark}[re]{Remark}
\newtheorem{Corollary}[thm]{Corollary}
\newtheorem{Definition}[thm]{Definition}
%% For proofs, please use the proof environment (the amsthm package is loaded by
% the MDPI class).

@@

% Authors (Add full first names)
\Author{Peter Bachant * and Martin Wosnik *}

% Affiliations / Addresses (Add [1] after \address if there is only one
% affiliation.)
\address[1]{%
 Center for Ocean Renewable Energy, University of New Hampshire, 24
Colovos Rd., \mbox{Durham, NH \hl{zip code}, USA}}

% Contact information of the corresponding author (Add [2] after \corres if
% there are more than one corresponding author.)
\corres[2]{\hspace{-1em} Correspondence: pxL3@unh.edu (P.B.); martin.wosnik@unh.edu (M.W.); \linebreak Tel.: \hl{(P.B.); (M.W.)}}

%\linenumbers

% Abstract (Less than 200 words)
\abstract{Experiments were performed with a large laboratory-scale high solidity
 cross-flow turbine to investigate Reynolds number effects on performance and
 wake characteristics and to establish scale thresholds for physical and
 numerical modeling of individual devices and arrays.~It was demonstrated that
 the performance of the cross-flow turbine becomes essentially $Re$-independent
 at a Reynolds number based on a turbine diameter of $Re_D \approx 10^6$ or an
 approximate average Reynolds number based on a blade chord of $Re_c \approx 2
 \times 10^5$.~A simple model that calculates the peak torque coefficient from static
 foil data and cross-flow turbine kinematics was shown to be a reasonable
 estimator for predicting the Reynolds number dependence of an actual cross-flow
 turbine operating under dynamic conditions. Mean velocity and turbulence
 measurements in the near-wake showed subtle differences over the range of $Re$
 investigated.~However, when transport terms for the streamwise momentum and mean
 kinetic energy were calculated, a similar $Re$ threshold was revealed. These
 results imply that physical model studies of cross-flow turbines should achieve
 $Re_D \sim 10^6$ to properly approximate both the performance and wake dynamics
 of full-scale devices and arrays.}

% Keywords: add 3 to 10 keywords
\keyword{\textls[-10]{Reynolds number; cross-flow turbine; turbine performance; marine
hydrokinetic energy}; wind energy; {VAWT}; scale model}

@@

Scaled physical models are often used in science and engineering to approximate
real-world systems. The principle of dynamic similarity allows for geometrically-scaled systems to be dynamically identical if certain nondimensional physical
parameters are matched. In the case of fluid systems, the most important
nondimensional parameter is often the Reynolds number, $Re$, which quantifies
the importance of inertial forces over viscous forces on the flow physics
\cite{Acheson1990}: \hl {$Re \equiv Ul/\nu$, } 
%please check the symbol ``\equiv '' and make sure it is correct
where $U$ and $l$ are characteristic
velocity and length scales, respectively, and $\nu$ is the fluid kinematic
viscosity. The~other common dynamical scale, for systems with a free
surface, is the Froude number \hl {$Fr \equiv U/\sqrt{gl}$,}
%please check the symbol ``\equiv '' and make sure it is correct
  where $g$ is the
gravitational acceleration. Matching both the Froude and Reynolds number is not
possible for a scaled model in a given fluid, which is illustrated by the
relation $Re = l^{3/2} g^{1/2} Fr / \nu$, since $Re$ scales linearly with $l$,
or the geometric scale. In general, the geometric scale of a physical model is
not necessarily the same as its dynamical scale. As such, hereafter, we will use
the word ``scale'' to refer to this dynamical scale or the Reynolds number and
assume the Froude number is small enough to be~ negligible.

With regards to wind and marine hydrokinetic (MHK) turbines, scaled physical
models are used to validate predictive numerical models, predict full-scale
performance of individual turbines and design or investigate arrays of devices.
Scaled models have the benefit of being significantly less expensive; however, a
key question is whether or not an acceptable scale mismatch exists. Similarly,~it is questionable whether numerical models validated with physical model data
obtained orders of magnitude away from the prototype scale should be considered
validated at all. An example of the errors that can result from the
extrapolation of small-scale experiments can be found in \cite{Baker1991}.

The performance and wake characteristics of cross-flow (vertical-axis) turbines
depend on turbine solidity, blade profile (lift/drag, dynamic stall at reduced
frequency of turbine rotation, symmetry, thickness, camber), blade pitch, number
of blades, strut drag and operational parameters, such as tip speed ratio, {
 and} on the Reynolds number \cite{Para2002}. Note that an average blade chord
Reynolds number, $Re_{c,\mathrm{avg}} \approx \lambda U_\infty c/ \nu$, can be
expressed in terms of tip speed ratio \hl {$\lambda \equiv \omega R/ U_\infty$, } 
%please check the symbol ``\equiv '' and make sure it is correct
where
$U_\infty$ is the free stream velocity, $c$ is the blade chord length, $\nu$ is
the fluid kinematic viscosity, $\omega$ is the rotor's angular velocity and $R$
is the rotor radius. The value of $\lambda$ at which a turbine reaches peak
performance in general decreases with turbine solidity $Nc/(\pi D)$
\cite{Templin1974}, which allows for the use of a simpler Reynolds number based
on turbine diameter $Re_D \equiv U_\infty D/\nu$.

Solidity often directly correlates with the chord-to-radius ratio $c/R$. Rotors
with $c/R \ge 0.1$ are considered to have high solidity \cite{Fiedler2009}, for which
so-called flow curvature or virtual camber effects become significant
\cite{Migliore1980}. These effects arise from the blade sections' circular
paths and can complicate the comparison with the behavior of an airfoil in a linear
flow.

Blackwell \emph{et al.} \cite{Blackwell1976} investigated the effects of Reynolds
number on the performance of a 2-m diameter Darrieus vertical-axis cross-flow
wind turbine with NACA 0012 blades. By varying wind tunnel speed, the turbine
was made to operate at an approximately constant blade chord Reynolds number $Re_c$
ranging from $1 \times 10^5$ to $3 \times 10^5$. In this regime, the turbine
power coefficient $C_P$ was shown to be sensitive to $Re_c$, with sensitivity
diminishing at the higher Reynolds numbers, especially for turbines of lower
solidity ($Nc/R$, where $N$ is the number of blades and $R$ is the turbine's
maximum radius). More recently, Polagye and Cavagnaro \cite{Polagye2013b}
observed significant Reynolds number sensitivity for a high solidity helical
cross-flow turbine, and Bravo \emph{et al.} \cite{Bravo2007} observed the power
coefficient of a straight-bladed VAWT to become Reynolds number independent at
$Re_c \approx 4 \times 10^5$. Bachant and Wosnik \cite{Bachant2014} observed
Reynolds number independence of the power coefficient at the optimal tip speed ratio for
a high solidity cross-flow turbine at $Re_D \approx 10^6$ or
$Re_{c,\mathrm{avg}} \approx 2 \times 10^5$.

The wake of a 2D cross-flow turbine in the dynamic stall regime has been
studied via laser Doppler velocimetry by Brochier \emph{et al.} \cite{Brochier1986} and
in 3D with particle image velocimetry (PIV) by Fujisawa and Shibuya
\cite{Fujisawa2001}. However, both of these studies were performed at very low
Reynolds numbers: $Re_D = 10^4$ and $10^3$, respectively. Tescione \emph{et al.}
\cite{Tescione2014} studied the wake of a vertical-axis wind turbine at the optimal
tip speed ratio in a wind tunnel using stereo PIV at an approximate blade chord
Reynolds number $Re_c = 1.7 \times 10^5$; very close to the $Re$-independence
criteria reported in \cite{Bachant2014}, though it was not confirmed if this was
indeed a $Re$-independent state. The two lower $Re$ experiments tended to focus
on the effects of dynamic stall, whereas the higher $Re$ experiment mostly
concerned the mean velocity and tip vortex effects. Whereas the value of these
studies was in elucidating the complex wake dynamics of cross-flow turbines,
they also motivated the more systematic investigation of scale effects
undertaken here.

Scale effects on axial-flow or horizontal-axis wind turbines are better
understood, and investigators have methods for compensating. Krogstad and
Adaramola \cite{Krogstad2012a} observed Reynolds number independence of the
performance of a 0.9-m diameter axial-flow turbine rotor at $Re_D \approx 5
\times 10^5$ in wind tunnel tests. Their turbine blades had an S826 profile
along their entire span, a profile chosen for its $Re$-independence. Walker \emph{et
al.} \cite{Walker2014} similarly chose a NACA 63-618 foil for their axial-flow
turbine tests in a towing tank, which were performed at a mid-span Reynolds
number $Re_c = 4 \times 10^5$. McTavish \emph{et al.} \cite{McTavish2013} showed how
the near-wake expansion for an axial-flow rotor is increased at higher Reynolds
numbers, concluding that physical models should be designed with
$Re$-independence in mind if they are to be run at low $Re$. It is uncertain,
however, if these methods work equally well for cross-flow turbines.

When designing or studying arrays, it is common to use very small
(geometrically) scaled \mbox{devices \cite{Chamorro2011, Chamorro2011b}}. It is
therefore important to realize the limitations of evaluating both the power
output of turbines and the wake recovery when the Reynolds number is very far from
full scale. Sometimes,~the models used are not turbines, but wake-generating objects,
e.g., porous disks, that are meant to replicate the wakes of real turbines
\cite{Goldenberg1983}. In this case, it is of interest to determine at what
Reynolds number one might be able to realistically study wake flows in an array
and also to evaluate the effectiveness of a wake generator. In other words, a
wake generator may do a fine job simulating a scaled turbine, but how well can
it simulate a full-scale device? Note that for a porous disk, the Reynolds
number of interest is diameter based, since it is the scaling of the far-wake
dynamics that~ matters.

Vertical-axis wind turbine array field experiments have revealed that improved
wake recovery allows for closer spacing when compared to conventional axial-flow
propeller-type \mbox{turbines \cite{Dabiri2011, Kinzel2012}}. It~was observed
experimentally that a high solidity vertical-axis cross-flow turbine's near-wake
produces a unique vertical mean velocity field, generated by blade tip vortex
shedding, the advection by which is the largest contributor to streamwise
momentum and energy transport or recovery \cite{Bachant2015-JoT}. In~this study,
we seek to replicate those same momentum and energy balance considerations at
multiple Reynolds numbers, to examine the implications on how scaled, \emph{i.e.}, low
Reynolds number experiments, may be used to study flows in turbine arrays.

@@

We are looking at how Reynolds number scaling affects both the blade loading,
\emph{i.e.}, turbine performance, and the near-wake. Typically, static airfoil data
show that the static stall angle increases with blade chord Reynolds number
$Re_c$ \cite{Jacobs1937}. A review of Reynolds number effects on airfoil
behavior is presented in \cite{Lissaman1983}. In general, airfoil
performance, often characterized by the profile's lift-to-drag ratio, is
enhanced as the boundary layer on the foil transitions to turbulence closer to
the leading edge, which enables it to advance further downstream against the
adverse pressure gradient on the suction side, delaying separation to higher
angles of attack. For smooth airfoils, this transition can cause a dramatic
increase in foil performance at a blade chord Reynolds number on the order of
\mbox{$10^5$ \cite{McMasters1980}}. Note~that there is a distinct lack of highly
reliable data for airfoils in this transitional regime and below. An evaluation
of the various databases relevant to cross-flow turbines is presented in
\cite{Bedon2014}.

\textls[-10]{Static foil performance does not tell the whole story for a cross-flow turbine.
The azimuthal, and} therefore, temporal, variation of $\alpha$ in a cross-flow
turbine implies dynamic loading, encountering \textls[-10]{dynamic stall for tip speed ratios
near and below those of maximum rotor torque \cite{Para2002}. Bousman
\cite{Bousman2000-evaluation}} states that dynamic stall is relatively
insensitive to Reynolds number for $Re=1.0 \times 10^5$ to $2.5 \times
10^5$, judging from measurements on a pitching VR-7 foil in a wind and water
tunnel, since the loading is dominated by vortex shedding. However, Singleton
and Yeager Jr. \cite{Singleton2000} state that the effect of Reynolds number on
dynamic stall remains an unsolved question.

Despite lower lift on the blades at lower $Re$, we expect stronger tip vortex
shedding \cite{Yoon2005}. As~mentioned previously, the dynamic stall vortex
shedding is not expected to be highly sensitive to Reynolds number, though a
larger separation bubble at lower $Re$ may induce higher levels of turbulence as
shed vortices become unstable and break down.

Chamorro \emph{et al.} \cite{Chamorro2012} showed that turbulence statistics in an AFT%please define
wake became $Re$-independent at $Re_D \approx 1 \times 10^5$ and that mean
velocity profiles became $Re$-independent slightly earlier at $Re_D \approx 5
\times 10^4$. Note that in this study, the tip chord Reynolds number was not
reported, but is estimated to be $Re_c \approx 4 \times 10^4$ at $Re_D=1 \times
10^5$. It is therefore our objective to observe similar scaling relationships
for a cross-flow turbine near-wake.

@@

Experiments were performed in a turbine test bed specifically designed for
cross-flow turbines. The test bed was integrated as part of the \hl {University of New Hampshire (UNH)} 
 tow tank,
which is a 36 m-long facility with a 3.66 m-wide and 2.44 m-deep cross-section.
The turbine model used in this study was the UNH Reference Vertical Axis Turbine
(RVAT), which was designed to be a generic case for numerical model testing,
similar to the Sandia National Labs/U.S. Department of Energy RM2%please define
 River Turbine
\cite{Neary2014}, but with a higher solidity or blade chord-to-radius ratio.

The UNH-RVAT turbine has three blades made from \hl {NACA} 0020 profiles
%please define NACA and add the information of Manufacturer, City, Country
  with a 0.14 m
chord. The~blades are mounted at mid-chord w.r.t. the turbine axis, and the
turbine has a height (blade span) of 1 m and a diameter of 1 m; \textit{cf}.
Figure~\ref{fig:turbine}. The blockage ratio produced by the rotor's frontal
area is 11\%, which we decided not to correct for, as reliable methods are
not yet agreed upon for \hl {CFTs} %please define
\cite{Cavagnaro2014}. As such, blockage should be
taken into account when using these data for model validation, \emph{i.e.}, the
experimental domain should be mimicked. The rotor has a relatively high solidity
$Nc/(\pi D) = 0.13$ and a large chord-to-radius ratio $c/R = 0.28$. A CAD model of
the turbine is available from \cite{Bachant2014-RVAT-CAD}.

\begin{figure}[H]
\centering

@@

\caption{\hl {University of New Hampshire (UNH)}-Reference Vertical Axis Turbine (RVAT) turbine model. Turbine blades and struts made from NACA 0020
 profiles with 0.140-m chord. Note that the upper and lower mounting flanges have
 been excluded, as these were included in the tare drag measurements.}

@@

The turbine was mounted in a frame constructed from NACA 0020 sections, shown in
Figure~\ref{fig:exp-setup}. The turbine shaft ran up through the water surface,
coupled to a Kollmorgen \hl {AKM}
%please define AKM and RPM
  permanent magnet servo \hl {motor (Manufacturer, City, Country)} with a 20:1 gearbox,
providing precise control over shaft \hl {RPM}. This servo was controlled by the tow
tank's main motion controller for high synchronization with the carriage motion,
thereby giving precise measurement and control of the tip speed ratio. An Interface
T8 200-Nm capacity rotary torque \hl {transducer (Manufacturer, City, Country)} was installed inline between the
servo and the turbine, and the servo was also mounted on a slewing ring bearing,
which allowed a redundant measurement of torque via an arm and load cell used to
counteract the turbine moment. The frame was mounted to the carriage via linear
guides, such that the total streamwise drag force was transferred to a pair of
Sentran \hl {ZB3 500-lbf} %please define/check the unit
 capacity S-beam \hl {load cells (Manufacturer, City, Country)}, providing the rotor drag
measurements, after a separately measured tare drag was subtracted in post-processing. Similarly, a tare torque was measured by rotating the turbine shaft
in air. Turbine angular and tow carriage linear position were measured using
quadrature encoder signals, with $10^5$ counts-per-rev for the turbine and 10 ${\mu}$m resolution for the carriage position. These~signals, along with
the torque and drag signals, were sampled at 2 kHz.

Wake velocity was measured using a Nortek Vectrino+ acoustic Doppler \hl {velocimeter
(ADV)  (Manufacturer, City, Country)}, which has an approximately 6-mm diameter sampling volume and sampled at
200 Hz. The probe was mounted on an automated positioning system, also
controlled by the tow tank's main motion controller. The ADV and \hl {DAQ (please define)} systems'
sampling times were synchronized by triggering the start of data acquisition via
a pulse sent from the motion controller. Additional details of the turbine and
experimental setup are described in \cite{Bachant2015-JoT}.

\begin{figure}[H]
 \centering

 \includegraphics[width=0.7\textwidth]{figures/exp-setup-photo}

 \includegraphics[width=0.7\textwidth]{figures/exp_setup_drawing}

 \caption{\hl {Experimental} setup: turbine test bed installed in the UNH tow tank.}
 %please mark (a) and (b) in the figure and label (a) and (b) in the figure caption.

@@

@@

\begin{figure}[H]
 \centering
